\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-05-05 analysis/paper/paper.Rmd
edit jesse's section

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- TRUE
```

@@

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras; the interested reader can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t) dt = E[dN_{i}(t)].$$

@@

Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual contributions of the form
$$P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
Therefore, we can define an estimating equation for $\theta$ as follows:
$$L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.$$
When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$. Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios. Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent. Nonetheless, Saarela [-@saarela2016case] showed that the logarithm of this estimating function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.

@@

# Case study 1--European Randomized Study of Prostate Cancer Screening

@@

# Case study 2--Bone-marrow transplant

@@

# Case study 3--SUPPORT Data

In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves. In the next case study, we explore an advanced feature.

For the third case study, we will demonstrate regularization during hazard estimation. Since we are using logistic regression, existing packages, like \pkg{glmnet} can be leveraged. To demonstrate this feature, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within 5 hospitals in the United States of America. Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice]. The imputed dataset is available as part of the \pkg{casebase} package. For more information please see the SUPPORT data documentation in the package.

we have a total of 9104 observations and 30 variables, two of which are SUPPORT day 3 physiology score (sps) and APACHE III day 3 physiology score (aps). These scores were developed for a period of 180 days, which makes our comparison over 5.56 years exploratory. Therefore, we will assess these two models before using either as our baseline, by comparing them to a full model using all variables. Also, our single imputation introduces bias in the data. As the goal here is purely demonstrative, this bias will be consistent with any of the following models. As such, their comparison is still valid, though any interpretation requires an awareness of said bias. <!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of covariates and response variables can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}. The original data was obtained from the Department of Biostatistics at Vanderbilt University [@harrell_2020].

@@

\caption{A description of each variable in the SUPPORT dataset.}

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
ca       & Yes                                           \\
         & No                                            \\
         & Metastatic                                    \\
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset.}
\label{tab:support2}
\end{table}

Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison. The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years. Therefore, We will examine three cox regression scenarios: sps only, aps only and all covariates excluding sps and aps. Next, we will compare each modeling using the Brier score [@graf1999ass]. The Brier score can demonstrate both discrimination and calibration in a single model [@graf1999ass] and provide a comparable metric between semi-parametric and parametric methods in the survival context. For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability. Over 5.56 years, \ref{fig:baselines} demonstrates that all the cox models have similar, if not identical average predictions. Figure \ref{fig:brier1} examines the brier scores as they change over time. As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.

```{r baselinePlotsPrep, eval=eval_cs3,echo=FALSE}
q4 <- qualitative_hcl(4, palette = "Dark 3")
completeData <- casebase::support

completeData$d.time <- completeData$d.time / 365

workingCompleteData <- model.matrix(death ~ . - d.time, data = completeData)[, -c(1)] # remove intercept

# Create u and xCox, which will be used when fitting Cox with glmnet.
newData <- workingCompleteData
x <- workingCompleteData
# x and y will be used to fit the casebase model under glmnet.
# x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y <- data.matrix(completeData[, c(2, 5)])

fullDataCox <- as.data.frame(cbind(as.numeric(y[, 2]), as.numeric(y[, 1]), x))
names(fullDataCox)[names(fullDataCox) == "V1"] <- "time"
names(fullDataCox)[names(fullDataCox) == "V2"] <- "event"

# here we specify train and test size
train_index <- sample(1:nrow(fullDataCox), 0.8 * nrow(fullDataCox))
test_index <- setdiff(1:nrow(fullDataCox), train_index)

train <- fullDataCox[train_index, ]
test <- fullDataCox[test_index, ]
coxSPS <- survival::coxph(Surv(time = time, event = event) ~ sps, data = train, x = TRUE)
abcoxSPS <- survival::survfit(coxSPS, newdata = test)

coxAPS <- survival::coxph(Surv(time = time, event = event) ~ aps, data = train, x = TRUE)
abcoxAPS <- survival::survfit(coxAPS, newdata = test)
coxfullForumla <- as.formula(paste("Surv(time = time, event = event) ~ age + sexmale + slos + dzgroupCHF + 
    dzgroupCirrhosis + `dzgroupColon Cancer` + dzgroupComa + 
    dzgroupCOPD + `dzgroupLung Cancer` + `dzgroupMOSF w/Malig` + 
    dzclassCancer + dzclassComa + `dzclassCOPD/CHF/Cirrhosis` + 
    num.co + edu + scoma + avtisst + raceasian + raceblack + 
    racehispanic + raceother + racewhite + hday + diabetes + 
    dementia + cano + cayes + meanbp + wblc + hrt + resp + temp + 
    pafi + alb + bili + crea + sod + ph + glucose + bun + urine + 
    adlp + adlsc "))
coxFull <- survival::coxph(coxfullForumla, data = train, x = TRUE)
abcoxFull <- survival::survfit(coxFull, newdata = test)
# library(prodlim)

KM <- survival::coxph(Surv(time = time, event = event) ~ 1, data = test)
abKM <- survival::survfit(KM)
# abKM <- prodlim(Hist(,event)~1,data=test) minimal difference
abKMcompare <- cbind(abKM$time[abKM$time <= 1631], 1 - abKM$surv[abKM$time <= 1631])
colnames(abKMcompare) <- c("Time", "KM")

baseCoxComparisonData <- cbind(abcoxSPS$time[abcoxSPS$time <= 1631], 1 - rowMeans(abcoxSPS$surv[abcoxSPS$time <= 1631, ]), 1 - rowMeans(abcoxAPS$surv[abcoxAPS$time <= 1631, ]), 1 - rowMeans(abcoxFull$surv[abcoxFull$time <= 1631, ]))

colnames(baseCoxComparisonData) <- c("Time", "SPS", "APS", "Full")
```

```{r baselines, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:baselines} Absolute risk curves estimated from hazard models. Red: APS only hazard was used to generate the absolute risk. Green: All covariates aside from APS and SPS. Blue: SPS only. Purple: unadjusted Kaplan-Meier curve.",echo=FALSE}
ggplot(as.data.frame(baseCoxComparisonData), aes(Time)) +
  geom_line(aes(y = APS, colour = "APS"), lwd = 1) +
  geom_line(aes(y = Full, col = "Full"), lwd = 1) +
  geom_line(aes(y = SPS, colour = "SPS"), lwd = 1) +
  geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 1) +
  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
  scale_color_manual(values = q4)
```

```{r brierpt1, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:brier1} Brier scores over survival time. Full model (Full) performs best, while the other two (APS only and SPS only) are comparable. Red: APS only model was used to generate the brier scores. Green: all covariates excluding APS and SPS. Blue: SPS only model.",echo=FALSE}

times <- sort(unique(test$time))
brierAPS <- Score(list("PredictionModel" = coxAPS), data = test, formula = Hist(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
brierSPS <- Score(list("PredictionModel" = coxSPS), data = test, formula = Hist(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)

coxFull$coefficients[which(is.na(coxFull$coefficients))] <- 0
brierFull <- Score(list("PredictionModel" = coxFull), data = test, formula = Surv(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)

brierPlotter <- rbind(
  data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"], times = times, model = "APS"),
  data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"], times = times, model = "Full"),
  data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"], times = times, model = "SPS")
)

ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) +
  geom_line() +
  xlab("Survival time (Years)") +
  ylab("Brier-Score") +
  labs(color = "Models") +
  ggtitle("Brier scores over time") +
  scale_color_manual(values = q4[1:3])
```

To get the absolute risk following regularized cox regression, we had to work around limitations in existing packages. To make use of the \pkg{survival} package, we first estimate the hazard coefficients using thecoxnet implementation for lasso penalization in \pkg{glmnet} [@regpathcox]. Next, we fit a cox model using the survival package with the selected variables only, from our regularized fit. Finally, we replaced the coefficients in the survival object with those from our glmnet fit, allowing us to make use of the absolute risk tools available in the \pkg{survival} package.

```{r coxNetCode,eval=FALSE,echo=TRUE}
# hazard for cox using glmnet
coxNet <- glmnet::cv.glmnet(
  x = x, y = y, family = "cox", alpha = 1,
  standardize = TRUE
)
# getting all non-zero covariates
nonzero_covariate_cox <- predict(coxNet,
  type = "nonzero",
  s = "lambda.1se"
)
# keeping coefficients for later
nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
# cutting dataset to only include nonzero covariates
cleanCoxData <- as.data.frame(cbind(y, x[, nonzero_covariate_cox]))
# create a skeleton object
coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ .,
  data = cleanCoxData, x = TRUE
)
# replace coefficients with regularized ones
coxNet$coefficients <- nonzero_coef_cox@x
```

For our penalized logistic regression model, we opted for the natural log of time, which assumes a Weibull distribution for our event. As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where *y* contains the time and event variables and *x* contains all covariates we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into dummy variables. Here, we used lasso penalization by setting alpha to 1, leading to our smooth hazard function.

```{r casebaseDisplayCode, eval=FALSE,echo=TRUE}
# regularized logistic regression to estimate hazard
cbFull <- casebase::fitSmoothHazard.fit(x, y,
  family = "glmnet",
  time = "d.time", event = "death",
  formula_time = ~ log(d.time), alpha = 1,
  ratio = 100, standardize = TRUE,
  penalty.factor = c(0, rep(1, ncol(x)))
)

# Estimating the absolute risk curve using the newData parameter.
abCbFull <- casebase::absoluteRisk(cbFull,
  time = times,
  newdata = newData,
  s = "lambda.1se",
  method = c("numerical")
)
```

```{r supportSetUp, echo = FALSE, eval = eval_cs3}

# remove apsSPS
train <- train[, -c(46, 47)]
test <- test[, -c(46, 47)]
# x and y will be used to fit the casebase model under glmnet.
y <- data.matrix(train[, c(2, 1)])
colnames(y) <- c("death", "d.time")
x <- data.matrix(train[, -c(2, 1)])
newData <- data.matrix(test[, -c(2, 1)])
```

```{r coxHazAbsolute, echo = FALSE,eval= eval_cs3 }
# Create u and xCox, which will be used when fitting Cox with glmnet.

u <- survival::Surv(time = as.numeric(y[, 2]), event = as.factor(y[, 1]))

xCox <- as.matrix(cbind(data.frame("(Intercept)" = rep(1, each = length(x[, 1]))), x))
colnames(xCox)[2] <- "(Intercept)"

# hazard for cox using glmnet
coxNet <- glmnet::cv.glmnet(x = x, y = u, family = "cox", alpha = 1, standardize = TRUE)
# convergence demonstrated in plot
# plot(coxNet)
# taking the coefficient estimates for later use
nonzero_covariate_cox <- predict(coxNet, type = "nonzero", s = "lambda.1se")
nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
# creating a new dataset that only contains the covariates chosen through glmnet.
cleanCoxData <- as.data.frame(cbind(d.time = as.numeric(y[, 2]), death = as.numeric(y[, 1]), x[, nonzero_covariate_cox$X1]))

# fitting a cox model using regular estimation, however we will not keep it.
# this is used more as an object place holder.
coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., data = cleanCoxData, x = TRUE)

# extract coefficients for lollipop plot later
estimatescoxNet <- setDT(as.data.frame(coef(coxNet)), keep.rownames = TRUE)[]

estimatescoxNet$Model <- "Penalized Cox"
colnames(estimatescoxNet) <- c("term", "estimate", "model")

# The coefficients of this object will be replaced with the estimates from coxNet.
# Doing so makes it so that everything is invalid aside from the coefficients.
# In this case, all we need to estimate the absolute risk is the coefficients.
# Std. error would be incorrect here, if we were to draw error bars.
coxNet$coefficients <- nonzero_coef_cox@x
# Fitting absolute risk curve for cox+glmnet. -1 to remove intercept

abcoxNet <- survival::survfit(coxNet, type = "breslow", newdata = as.data.frame(newData[, nonzero_covariate_cox$X1]))

brierCoxNet <- Score(list("PredictionModel" = coxNet), data = as.data.frame(cbind(test[, c(1, 2)], newData[, nonzero_covariate_cox$X1])), formula = Hist(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)

abCN <- as.data.frame(cbind(abcoxNet$time, 1 - rowMeans(abcoxNet$surv)))
colnames(abCN) <- c("Time", "Absolute risk")
abCN$Model <- "Penalized Cox"
abCN <- abCN[abCN$Time <= max(abKMcompare[, 1]), ]
# plot(abCN[,1],1-abCN[,2])
```

```{r cbHazard}, echo = FALSE, eval = eval_cs3}

cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE, penalty.factor = c(0, rep(1, ncol(x))))
```

```{r CBabsolute, echo = FALSE, eval = eval_cs3}
# Estimating the absolute risk curve using the newData parameter.
abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
```

```{r cbAbProcessing,echo=FALSE,eval=eval_cs3}
abCB <- data.frame(abCbFull[, 1], rowMeans(abCbFull[, -c(1)]))

colnames(abCB) <- c("Time", "Absolute risk")
abCB$Model <- "Penalized logistic"

@@

```{r coefplots,echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (green), penalized Cox (red), penalized logisitic (blue). If a coefficient is estimated to be 0, in any model, it is removed from the plot for simplicity."}

library(dotwhisker)
library(broom)
library(dplyr)

estimatescb <- setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), keep.rownames = TRUE)[]
# estimatescb$std.error<-0
estimatescb$Model <- "Penalized logistic"
colnames(estimatescb) <- c("term", "estimate", "model")

>>>>

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- FALSE
eval_cs2 <- FALSE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
```

@@

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras; the interested reader can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modelling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t) dt = E[dN_{i}(t)].$$

@@

Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual contributions of the form
$$P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
Therefore, we can define an estimating equation for $\theta$ as follows:
$$L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.$$
When a logarithmic link function is used for modelling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$. Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios. Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent. Nonetheless, Saarela [-@saarela2016case] showed that the logarithm of this estimating function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.

@@

# Case study 1---European Randomized Study of Prostate Cancer Screening

@@

# Case study 2---Bone-marrow transplant

@@

# Case study 3---SUPPORT Data

In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves. <!--In the next two case studies, we explore more advanced features.--> For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function. To do this, we simply replace logistic regression with a penalized counterpart. In \pkg{casebase}, we achieve this by using the \pkg{glmnet} package [@friedman2010jss].

To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support]. The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill. The original data is available online from the Department of Biostatistics at Vanderbilt University [@harrell_2020]. The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package. For more information about this dataset, the reader is encouraged to look at the documentation in our package. A description of the variables can be found in Table \ref{tab:support1}, and a breakdown of each categorical variable appears in Table \ref{tab:support2}.

@@

\caption{Description of each variable in the SUPPORT dataset.}

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
ca       & Yes                                           \\
         & No                                            \\
         & Metastatic                                    \\
\end{tabular}
\caption{Description of each level within each categorical variable in the SUPPORT dataset.}
\label{tab:support2}
\end{table}

<!--Two variables are SUPPORT day 3 physiology score (sps) and APACHE III day 3 physiology score (aps). These scores were developed for a period of 180 days, which makes our comparison over 5.56 years exploratory. Therefore, we will assess these two models before using either as our baseline, by comparing them to a full model using all variables. Also, our single imputation introduces bias in the data. As the goal here is purely demonstrative, this bias will be consistent with any of the following models. As such, their comparison is still valid, though any interpretation requires an awareness of said bias. The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.-->

In the original study, the authors developed two scores for predicting the outcome. These scores are available as part of the data: SUPPORT day 3 physiology score (\code{sps}) and APACHE III day 3 physiology score (\code{aps}). Using these scores, we will explore three Cox regression scenarios as a baseline comparison for \pkg{casebase}: \code{sps} only, \code{aps} only, and all covariates excluding \code{sps} and \code{aps}. We will compare the three models using a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass]. The Brier score can be used to assess both discrimination and calibration in a single model. Moreover, it provides a metric that can be used to compare parametric and semi-parametric survival models. To account for overfitting, we will use 95% of the observations to train our models, while the remaining 5% will be used to compute absolute risk curves and the Brier scores.

```{r supportData, eval = eval_cs3}
data(support)
# Change time to years
support$d.time <- support$d.time/365.25

# Split into test and train
train_index <- sample(nrow(support), 0.95*nrow(support))
test_index <- setdiff(1:nrow(support), train_index)

train <- support[train_index,]
test <- support[test_index,]
```

In Figure\ref{fig:baselines}, we can see that all Cox models have similar average survival curves. On the other hand, Figure \ref{fig:brier1} shows that the full model has the lowest Brier score overall. For this reason, we will use this model in our comparison between regularized Cox regression, Kaplan-Meier estimation, and the approach from \pkg{casebase}.

<!--Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison. The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years. Therefore, we will examine three Cox regression scenarios: sps only, aps only and all covariates excluding sps and aps. Next, we will compare each modeling using the Brier score [@graf1999ass]. The Brier score can demonstrate both discrimination and calibration in a single model [@graf1999ass] and provide a comparable metric between semi-parametric and parametric methods in the survival context. For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability. Over 5.56 years, \ref{fig:baselines} demonstrates that all the cox models have similar, if not identical average predictions. Figure \ref{fig:brier1} examines the brier scores as they change over time. As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.-->

```{r baselinePlotsPrep, eval=eval_cs3, echo = FALSE}
# Define colour palette for this section
q4 <- qualitative_hcl(4, palette = "Dark 3")

# 1. Cox with only sps
coxSPS <- survival::coxph(Surv(time = d.time, event = death) ~ sps, 
                          data = train, x = TRUE)
abcoxSPS <- survival::survfit(coxSPS, newdata = test)
# 2. Cox with only aps
coxAPS <- survival::coxph(Surv(time = d.time, event = death) ~ aps, 
                          data = train, x = TRUE)
abcoxAPS <- survival::survfit(coxAPS, newdata = test)
# 3. Cox with everything but sps and aps
coxFull <- survival::coxph(Surv(time = d.time, event = death) ~ . - aps - sps, 
                           data = train, x = TRUE)
abcoxFull <- survival::survfit(coxFull, newdata = test)
# 4. Kaplan-Meier
KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test)
abKM <- survival::survfit(KM)
abKMcompare <- data.frame(
  Time = abKM$time,
  KM = 1 - abKM$surv
)

# Combine all CI estimates from Cox models
baseCoxComparisonData <- data.frame(
  Time = abcoxSPS$time,
  SPS = 1 - rowMeans(abcoxSPS$surv),
  APS = 1 - rowMeans(abcoxAPS$surv),
  Full = 1 - rowMeans(abcoxFull$surv)
)
```

```{r baselines, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:baselines} Cumulative Incidence curves estimated using Cox regression and Kaplan-Meier (K-M)."}
ggplot(baseCoxComparisonData, aes(Time)) +
  geom_line(aes(y = APS, colour = "APS"), lwd = 1) +
  geom_line(aes(y = Full, col = "Full"), lwd = 1) +
  geom_line(aes(y = SPS, colour = "SPS"), lwd = 1) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M"))
```

```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Time-dependent Brier scores."}
# Extract time points
times <- sort(unique(test$d.time))
# Compute Brier scores
brierAPS <- Score(list("PredictionModel" = coxAPS), data = test, 
                  formula = Hist(d.time, death != 0) ~ 1, summary = "ipa", 
                  se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
brierSPS <- Score(list("PredictionModel" = coxSPS), data = test, 
                  formula = Hist(d.time, death != 0) ~ 1, summary = "ipa", 
                  se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)

# First fix NA coefficients, then compute Brier score
coxFull$coefficients[which(is.na(coxFull$coefficients))] <- 0
brierFull <- Score(list("PredictionModel" = coxFull), data = test, 
                   formula = Surv(d.time, death != 0) ~ 1, summary = "ipa", 
                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)

# Combine all results for plotting
brierPlotter <- rbind(
  data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"], 
             times = times, model = "APS"),
  data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"], 
             times = times, model = "Full"),
  data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"], 
             times = times, model = "SPS")
)

ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) +
  geom_line() +
  xlab("Survival time (Years)") +
  ylab("Brier score") +
  labs(color = "Models") +
  # ggtitle("Brier scores over time") +
  scale_color_manual(values = q4[1:3])
```

<!--To get the absolute risk following regularized cox regression, we had to work around limitations in existing packages. To make use of the \pkg{survival} package, we first estimate the hazard coefficients using thecoxnet implementation for lasso penalization in \pkg{glmnet} [@regpathcox]. Next, we fit a cox model using the survival package with the selected variables only, from our regularized fit. Finally, we replaced the coefficients in the survival object with those from our glmnet fit, allowing us to make use of the absolute risk tools available in the \pkg{survival} package.-->

For our penalized logistic regression model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution. For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the function \code{fitSmoothHazard}. In order to use this interfact, we first create a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all remaining covariates we would like to include in the model space. We used lasso penalization by setting \code{alpha = 1}. Moreover, we use the argument \code{penalty.factor} to force the time variable into the final model.

```{r supportCB_fit, eval=eval_cs3, echo=TRUE, cache=TRUE}
x <- model.matrix(death ~ . - d.time - aps - sps, 
                  data = train)[, -c(1)] # Remove intercept
y <- data.matrix(subset(train, select = c(d.time, death)))

# Regularized logistic regression to estimate hazard
cbFull <- casebase::fitSmoothHazard.fit(x, y,
  family = "glmnet",
  time = "d.time", event = "death",
  formula_time = ~ log(d.time), alpha = 1,
  ratio = 10, standardize = TRUE,
  penalty.factor = c(0, rep(1, ncol(x)))
)
```

Next, we can use the object \code{cbFull} to estimate the absolute risk curve on the test data. We first create a matrix \code{newx} containing the relevant variables. Next, using \code{times}, we specify the time points at which the cumulative incidence will be computed. Finally, using the parameter \code{s = "lambda.1se"}, we specify at which value of the tuning parameter we want to retrieve the coefficient estimates.

```{r supportCB_abs, eval=eval_cs3, echo=TRUE, cache=TRUE}
# Estimating the absolute risk curve using the newdata parameter
newx <- model.matrix(death ~ . - d.time - aps - sps, 
                  data = test)[, -c(1)]
times <- sort(unique(test$d.time))

abCbFull <- casebase::absoluteRisk(cbFull,
  time = times,
  newdata = newx,
  s = "lambda.1se",
  method = "numerical"
)
```

```{r coxHazAbsolute, echo = FALSE, eval = eval_cs3}
# Create survival object for fitting Coxnet
u <- with(train, survival::Surv(time = d.time, event = death))
coxNet <- glmnet::cv.glmnet(x = x, y = u, family = "cox", alpha = 1, standardize = TRUE)

# Taking the coefficient estimates for later use
nonzero_covariate_cox <- predict(coxNet, type = "nonzero", s = "lambda.1se")
nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
# Creating a new dataset that only contains the covariates chosen through glmnet
cleanCoxData <- as.data.frame(cbind(y, x[, nonzero_covariate_cox$X1]))

# Fitting a cox model using regular estimation, however we will not keep it.
# this is used more as an object place holder.
coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
                          data = cleanCoxData, x = TRUE)

# The coefficients of this object will be replaced with the estimates from the
# original coxNet. Doing so makes it so that everything is invalid aside from
# the coefficients. In this case, all we need to estimate the absolute risk is
# the coefficients. Std. error would be incorrect here, if we were to draw error
# bars.
coxNet_coefnames <- names(coxNet$coefficients)
coxNet$coefficients <- nonzero_coef_cox@x
names(coxNet$coefficients) <- coxNet_coefnames

# Fitting absolute risk curve for cox+glmnet
abcoxNet <- survival::survfit(coxNet, type = "breslow", 
                              newdata = as.data.frame(newx[, nonzero_covariate_cox$X1]))

# Combine CI estimates from both penalized methods
abCN <- data.frame(
  Time = abcoxNet$time,
  absRisk = 1 - rowMeans(abcoxNet$surv)
)
abCN$Model <- "Penalized Cox"

abCB <- data.frame(Time = abCbFull[, 1],
                   absRisk = rowMeans(abCbFull[, -c(1)]))
abCB$Model <- "Penalized logistic"

@@

Our first comparison plot examines the covariates that were selected between all three hazard functions. Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of coefficients with values greater than 0 are displayed in \ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have larger estimated coefficients. This is expected, as there are no restrictions on the size of coefficients. The smallest coefficient estimates are found in penalized logistic model, aside from cancer state (cano).

```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models. If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity."}
library(dotwhisker)
library(broom)
library(dplyr)

# extract coefficients for lollipop plot
estimatescoxNet <- setDT(as.data.frame(coef(coxNet)), 
                         keep.rownames = TRUE)
